### 2022-04-26 misc/expo/tensorrank.tex
Tensorrank typos

\medskip\boldlabel The cap set problem. The partition rank was introduced to study the cap-set problem,
and here we shall sketch how it applies. A {\it cap set} is a subset $A\subseteq \FF_3^n$ such that for
every triple $(x,y,z)\in A$ of pairwise distinct elements, $x+y+z\ne 0$. In 1984, T.~C.~Brown
and J.~P.~Buhler showed that, loosely speaking, cap sets have zero density.

@@

Given $A\subseteq \FF_3^n$, let $T : V^3 \to \FF_3$, where $V = {\FF_3}^{\FF_3^n}$, be given by
$$f(e_a,e_b,e_c = \cases{1, & if $a+b+c=0$;\cr 0, & otherwise,}$$
for basis vectors $e_v$ and extended to all other vectors by linearity. (The function $e_v$ has
$e_v(v) = 1$ and $e_v(x)=0$ for all $x\ne v$.)
Now for a tensor $T : (\FF^X)^d \to \FF$, we say that a set $A\subseteq X$ is an {\it independent
set in $T$} if for all $i_1,\ldots,i_d\in A$, the condition that the coefficient $T_{i_1,\ldots,i_d}$
be nonzero is equivalent to $i_1=\cdots=i_d$.
We then give an upper bound on the size of a cap set by proving that
\medskip
\item{i)} if $A$ contains no nontrivial solutions to $x+y+z=0$, then $A$ is an independent set in $T$;
\smallskip
\item{ii)} if $A$ is an independent set in $T$ then $\prk(T)\ge |A|$; and
\smallskip
\item{iii)} the partition rank of $T$ is low.
\medskip
In these notes, we aim to show that this general strategy may be performed with the partition rank
replaced by something called the analytic rank.

@@

\proof We expand
$$\bias(B) = \ex_{x\in \FF^n, y\in \FF_m} \chi\bigl(B(x,y)\bigr)
= \ex_{y\in \FF^m} \one_{W_1(y) = \cdots = W_n(y) = 0} \cdot \chi\bigl(W_0(y)\bigr)$$
and by the triangle inequality,
$$\bigl|\bias(B)\bigr| \le \ex_y \one_{W_1(y) = \cdots = W_n(y) = 0} = \bias(A).\noskipslug$$

@@

Before proving subadditivity, we introduce another bit of notation. For $I\subseteq [d]$,
let
$$T_I(\x,\y) = T(\x_I, y_{I^c}) = T(z_1, \ldots, z_d),$$
where $z_i = x_i$ if $i\in I$ and $z_i = y_i$ if $i\notin I$. After expanding out the definition
of multilinearity, we see that $T(\x+\y)$ decomposes as
$$ T(\x+\y) = \sum_{I\subseteq[d]} T_I(\x,\y).$$

@@

\proof It is enough to show that
$$\bias(T+S) \ge \bias(T)\bias(S).$$
We express
$$\eqalign{
\bias(T)\bias(S) &= \Bigl(\ex_{\x\in V^d} \chi\bigl(T(\x)\bigr)\Bigr)
                    \Bigl(\ex_{\y\in V^d} \chi\bigl(T(\y)\bigr)\Bigr)\cr
&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\y)\bigr) \cr
&= \ex_{\x\in X} \ex_{\y\in Y} \chi\bigl( T(\x) + S(\x+\y)\bigr) \cr
&= \bias\bigl( T(\x) + S(\x+\y)\bigr), \cr
}$$
and by our earlier decomposition the tensor product of a sum, this gives us
$$\bias(T)\bias(S)
= \bias \Bigl( T(\x) + \sum_{I\subseteq [d]} S_I(\x,\y) \Bigr)
= \bias \Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x,\y) \Bigr), $$
Setting $\y = \b\in V^d$ to maximise the norm of this right-hand side, we have
$$\bias(T)\bias(S) \le \Bigl| \bias\Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x, \b)\Bigr)\Bigr|.$$
Here $S_I(\x,\b)$ is an order-$|I|$ tensor in the variables $\x_I$. Thus we may apply
the previous lemma with
$$R_{[d]}(\x) = (T+S)(\x)\qquad\hbox{and}\qquad R_I(\x_I) = S_I(\x,\b)\ \hbox{for}\ I\ne [d]$$
to get $\bias(T)\bias(S) \le \bias(T+S)$.\slug

@@

\proclaim Theorem \advthm. Let $T : (\FF^n)^d\to \FF$ be an order-$d$ tensor. Assume that
$A\subseteq [n]$ is an independent set in $T$. Then $\ark(T)\ge c|A|$ where
$$c = -\log_{|\FF|} \biggl( 1-\biggl( 1-{1\over y}\biggr)^{d-1}\biggr),$$
so that $c\ge 2^{-d}$ and $c\ge 1-\log(d-1)/\log |\FF|$.

\proof Let $S : (F^A)^d\to \FF$ be the restriction of $T$ to $\FF^A$.
By the previous lemma, $\ark(S)$ is a lower bound on $\ark(T)$, so we shall prove the theorem
by expressing $\ark(S)$ as a function of $|A|$. We have
$$\bias(A) = \ex_{x_1,\ldots,x_d\in \FF^A} \chi\Bigl( \sum_{i\in A} c_i \prod_{j\in [d]} x_{i,j}\Bigr),$$
where, since $A$ is an independent set, $c_i\ne 0$ for all $i\in A$. But we also know that
$$\bias(S) =
\pr_{x_2,\ldots,x_d\in \FF^A} \Bigl\{ \bigcap_{i\in A} \prod_{2\le j\le d} \{x_{i,j} = 0\}\Bigr\},$$
and this is easily seen to be
$$\bias(S) = \biggl( 1-\biggl( 1-{1\over |\FF|}\biggr)^{d-1} \biggr)^{|A|}.$$
Taking the logarithm of both sides, we see that $\ark(S) = c|A|$, where $c$ is as in the theorem
statement.

>>>>

\medskip\boldlabel The cap set problem. The partition rank was introduced to study the cap-set problem,
and here we shall sketch how it applies. A {\it cap set} is a subset $A\subseteq \FF_3^n$ such that for
every triple $(x,y,z)\in A$ of pairwise distinct elements, $x+y+z\ne 0$. It was shown by T.~C.~Brown
and J.~P.~Buhler that, loosely speaking, cap sets have zero density.

@@

Given $A\subseteq \FF_3^n$, let $T : V^3 \to \FF_3$, where $V = {\FF_3}^{\FF_3^n}$, be given by
$$T(e_a,e_b,e_c) = \cases{1, & if $a+b+c=0$;\cr 0, & otherwise,}$$
for basis vectors $e_v$ and extended to all other vectors by linearity. (The function $e_v$ has
$e_v(v) = 1$ and $e_v(x)=0$ for all $x\ne v$.)
Now for a tensor $T : (\FF^X)^d \to \FF$, we say that a set $A\subseteq X$ is an {\it independent
set in $T$} if for all $i_1,\ldots,i_d\in A$, the condition that the coefficient $T_{i_1,\ldots,i_d}$
be nonzero is equivalent to $i_1=\cdots=i_d$.
We then give an upper bound on the size of a cap set by proving that
\medskip
\item{i)} if $A$ contains no nontrivial solutions to $x+y+z=0$, then $A$ is an independent set in $T$;
\smallskip
\item{ii)} if $A$ is an independent set in $T$ then $\prk(T)\ge |A|$; and
\smallskip
\item{iii)} the partition rank of $T$ is low.
\medskip
In these notes, we aim to show that this general strategy may be performed with the partition rank
replaced by something called the analytic rank.

@@

\proof We expand
$$\bias(B) = \ex_{x\in \FF^n, y\in \FF^m} \chi\bigl(B(x,y)\bigr)
= \ex_{y\in \FF^m} \indic{W_1(y) = \cdots = W_n(y) = 0} \cdot \chi\bigl(W_0(y)\bigr)$$
and by the triangle inequality,
$$\bigl|\bias(B)\bigr| \le \ex_y \indic{W_1(y) = \cdots = W_n(y) = 0} = \bias(A).\noskipslug$$

@@

Before proving subadditivity, we introduce another bit of notation. For $I\subseteq [d]$,
let
$$T_I(\x,\y) = T(\x_I, \y_{I^c}) = T(z_1, \ldots, z_d),$$
where $z_i = x_i$ if $i\in I$ and $z_i = y_i$ if $i\notin I$. After expanding out the definition
of multilinearity, we see that $T(\x+\y)$ decomposes as
$$ T(\x+\y) = \sum_{I\subseteq[d]} T_I(\x,\y).$$

@@

\proof It is enough to show that
$$\bias(T+S) \ge \bias(T)\bias(S).$$
We express
$$\eqalign{
\bias(T)\bias(S) &= \Bigl(\ex_{\x\in V^d} \chi\bigl(T(\x)\bigr)\Bigr)
                    \Bigl(\ex_{\y\in V^d} \chi\bigl(T(\y)\bigr)\Bigr)\cr
&= \ex_{\x\in V^d} \ex_{\y\in V^d} \chi\bigl( T(\x) + S(\y)\bigr) \cr
&= \ex_{\x\in V^d} \ex_{\y\in V^d} \chi\bigl( T(\x) + S(\x+\y)\bigr) \cr
&= \bias\bigl( T(\x) + S(\x+\y)\bigr), \cr
}$$
and by our earlier decomposition the tensor product of a sum, this gives us
$$\bias(T)\bias(S)
= \bias \Bigl( T(\x) + \sum_{I\subseteq [d]} S_I(\x,\y) \Bigr)
= \bias \Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x,\y) \Bigr), $$
Setting $\y = \b\in V^d$ to maximise the norm of this right-hand side, we have
$$\bias(T)\bias(S) \le \Bigl| \bias\Bigl( (T+S)(\x) + \sum_{I\ne [d]} S_I(\x, \b)\Bigr)\Bigr|.$$
Here $S_I(\x,\b)$ is an order-$|I|$ tensor in the variables $\x_I$. Thus we may apply
the previous lemma with
$$R_{[d]}(\x) = (T+S)(\x)\qquad\hbox{and}\qquad R_I(\x_I) = S_I(\x,\b)\ \hbox{for}\ I\ne [d]$$
to get $\bias(T)\bias(S) \le \bias(T+S)$.\slug

@@

\proclaim Theorem \advthm. Let $T : (\FF^n)^d\to \FF$ be an order-$d$ tensor. Assume that
$A\subseteq [n]$ is an independent set in $T$. Then $\ark(T)\ge c|A|$ where
$$c = -\log_{|\FF|} \biggl( 1-\biggl( 1-{1\over |\FF|}\biggr)^{d-1}\biggr),$$
so that $c\ge 2^{-d}$ and $c\ge 1-\log(d-1)/\log |\FF|$.

\proof Let $S : (F^A)^d\to \FF$ be the restriction of $T$ to $\FF^A$.
By the previous lemma, $\ark(S)$ is a lower bound on $\ark(T)$, so we shall prove the theorem
by expressing $\ark(S)$ as a function of $|A|$. We have
$$\bias(S) = \ex_{x_1,\ldots,x_d\in \FF^A} \chi\Bigl( \sum_{i\in A} c_i \prod_{j\in [d]} x_{i,j}\Bigr),$$
where, since $A$ is an independent set, $c_i\ne 0$ for all $i\in A$. But we also know that
$$\bias(S) =
\pr_{x_2,\ldots,x_d\in \FF^A} \Bigl\{ \bigcap_{i\in A} \{x_{i,2}\cdots x_{i,d} = 0\}\Bigr\},$$
and this is easily seen to be
$$\bias(S) = \biggl( 1-\biggl( 1-{1\over |\FF|}\biggr)^{d-1} \biggr)^{|A|}.$$
Taking the logarithm of both sides, we see that $\ark(S) = c|A|$, where $c$ is as in the theorem
statement.
